Staged Abstract Interpreters

### 2019-09-16 paper/30_stageconc.tex
polish

Lightweight Modular Staging (LMS) \cite{DBLP:conf/gpce/RompfO10} is a
multi-stage programming framework implemented as a Scala library. LMS enables
runtime code generation in a type-safe manner. Different from the approach of
MetaML/MetaOCaml \cite{DBLP:conf/flops/Kiselyov14,
DBLP:conf/gpce/CalcagnoTHL03} which uses syntactic quotations, LMS
distinguishes binding-times based on types and preserves the evaluation order
via an internal ANF transformation of the staged program.
To achieve typed-based MSP, LMS provides a type constructor @Rep[T]@, where @T@
can be an arbitrary type, indicating that a value of type @T@ will be known at
a future stage.  Operations acting on @Rep[T]@ expressions will be residualized
as generated code.
A classic example for introducing multi-stage programming is the power function
that computes $b^x$. Usually, it is implemented as a recursive function:
\begin{lstlisting}
  def power(b: Int, x: Int): Int = if (x == 0) 1 else b * power(b, x - 1)
\end{lstlisting}

@@

Readers may notice that the conversion between unstaged monads and staged monads
is merely changing the type of unstaged data to staged data, which in fact can
be achieved without modifying the implementation of monads. This is true so far,
but as we will see, it is not as straightforward for the nondeterminism monad
(@SetT@) in the abstract interpreter, because not only the elements in the set are
staged, but the whole set is also staged, and we have no knowledge about how
many elements are in the set. Therefore, any traversals over the set has to be
staged.  In this section, we explicitly distinguish the two versions of monadic
interfaces. Later, we will instantiate the staged set monad by manually fusing
the fragment of the monad stack inside @SetT@.

\paragraph{Primitive Operations} Most of the primitive operations can be easily
translated to their staged versions -- we just need to change the types.  As we
mentioned before, we will illustrate in detail how functions and applications
are handled.  One problem here is what should we do for $\lambda$-terms when
staging them? We cannot directly create a next-stage defunctionalized closure
for it, because that means we still need to interpret over these closures at the
next stage and the interpretation overhead has not been eliminated. 
The solution is to compile the $\lambda$-term with its environment by calling
the evaluator at the current stage. The result is generating a next-stage
Scala function. The evaluator is passed as @ev@ to the method @close@.
The following code implements this idea:
\begin{lstlisting}
  type ValSt = (Value, Store)
  def emit_compiled_clo(f: (Rep[Value], Rep[Store]) => Rep[ValSt], λ: Lam, ρ: Exp[Env]): Rep[Value]
  def close(ev: Expr => Ans)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value], Rep[Store]) => Rep[ValSt] = { case (v: Rep[Value], σ: Rep[Store]) =>
        val α = alloc(σ, x)
        ev(e)(ρ + (unit(x) → α))(σ + (α → v))
    }; emit_compiled_clo(f, λ, ρ)
  }
\end{lstlisting}

>>>>

Lightweight Modular Staging (LMS) \cite{DBLP:conf/gpce/RompfO10} is a
multi-stage programming framework implemented as a Scala library. LMS enables
runtime code generation in a type-safe manner. Different from the approach of
MetaML/MetaOCaml \cite{DBLP:conf/flops/Kiselyov14,
DBLP:conf/gpce/CalcagnoTHL03} which uses syntactic quotations, LMS
distinguishes binding-times based on types and preserves the evaluation order
via an internal ANF transformation of the staged program.
To achieve typed-based MSP, LMS provides a type constructor @Rep[T]@, where @T@
can be an arbitrary type, indicating that a value of type @T@ will be known at
a future stage.  Operations acting on @Rep[T]@ expressions will be residualized
as generated code.
A classic example of multi-stage programming is the power function
that computes $b^x$. Usually, it is implemented as a recursive function:
\begin{lstlisting}
  def power(b: Int, x: Int): Int = if (x == 0) 1 else b * power(b, x - 1)
\end{lstlisting}

@@

Readers may notice that the conversion between unstaged monads and staged monads
is merely changing the type of unstaged data to staged data, which in fact can
be achieved without modifying the implementation of monads. This is true so far,
but as we will see, it is not as straightforward for the nondeterminism monad
(@SetT@) in the abstract interpreter, because not only the elements in the set are
staged, but the whole set is also staged, and we have no knowledge about how
many elements are in the set. Therefore, any traversals over the set has to be
staged.  In this section, we explicitly distinguish the two versions of monadic
interfaces. Later, we will instantiate the staged set monad by manually fusing
the fragment of the monad stack inside the @SetT@ transformer.

\paragraph{Primitive Operations} Most of the primitive operations can be easily
translated to their staged versions -- we just need to change the types.  As we
mentioned before, we will illustrate in detail how functions and applications
are handled.  One problem here is what should we do for $\lambda$-terms when
staging them? We cannot directly create a next-stage defunctionalized closure
for it, because that means we still need to interpret over these closures at the
next stage and the interpretation overhead has not been eliminated. 
The solution is to compile the $\lambda$-term with its environment by calling
the evaluator at the current stage. The result is generating a next-stage
Scala function. The evaluator is already passed as @ev@ to the method @close@.
The following code implements this idea:
\begin{lstlisting}
  type ValSt = (Value, Store)
  def emit_compiled_clo(f: (Rep[Value], Rep[Store]) => Rep[ValSt], λ: Lam, ρ: Exp[Env]): Rep[Value]
  def close(ev: Expr => Ans)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value], Rep[Store]) => Rep[ValSt] = { case (v: Rep[Value], σ: Rep[Store]) =>
        val α = alloc(σ, x)
        ev(e)(ρ + (unit(x) → α))(σ + (α → v))
    }; emit_compiled_clo(f, λ, ρ)
  }
\end{lstlisting}

### 2019-09-16 paper/40_absinterp.tex
polish

In this section, we instantiate @SPLattice@ with the flat binding-time
type @R[T] = T@ (i.e., @NoRep@); the result is the trait @Lattice[S]@.
An example of the lattices we use in the rest of the paper is the power set
abstract domain (shown in Figure \ref{fig:powerset}). Two power sets are
ordered by the subset relation. We use set union to compute their
join, and set intersection to compute their meet.  The bottom element of a
power set is the empty set, and we do not have a representation of the top element for
power set.  Other lattices used in the paper, such as products and maps, can be
implemented similarly or by lifting the existing lattices element-wise or
point-wise.  Non-relational numerical abstract domains such intervals can also
be implemented in a stage-polymorphic way.

@@

In this section, we follow the abstracting definitional interpreters
(ADI) approach \cite{DBLP:journals/pacmpl/DaraisLNH17} to refactor our monadic
concrete interpreter to a monadic abstract interpreter.

@@

Similar to the concrete scenario, we sketch the @flatMap@ implementation of the
@SetT@ transformer and omit the rest. The field @run@ encapsulated by the monad
is of type @M[SetT[A]]@, where @M[_]@ is the inner monad, and @A@ is the element
type of the set. We first apply @flatMap@ on the inner monad to obtain the
set @s@. Then, we use the function @f@ to transform every element of type @A@
into a monadic value of type @Set[M, B]@. Finally, we fold all the transformed
values into a single monadic value of type @Set[M, B]@ by appending all of them.
The initial value of the fold is an empty @SetT@.
\begin{lstlisting}
  case class SetT[M[_]: Monad, A](run: M[Set[A]]) {
    def flatMap[B](f: A => SetT[M, B]): SetT[M, B] =
      SetT(Monad[M].flatMap(run) { (s: Set[A]) =>
        s.foldLeft(SetT.empty[M, B])((acc, a) => acc ++ f(a)).run
      }); ... }
\end{lstlisting}

@@

\paragraph{Primitive Operations} The primitive operations are changed according
to the new monad stack scheme and value domains. One of the notable changes is
that the store update operator merges the new values with existing values.  As
we mentioned before, sometimes we have to explore both of the branches for
conditionals: in method @br0@, we combine the results using the @mplus@
operation from @MonadPlus@, which requires the value domains to be
join-semilattices (in our case, @Set@ and @Map@).
\begin{lstlisting}
  def set_store(αv: (Addr, Value)): AnsM[Unit] = liftM(StateTMonad.mod(σ => σ ⊔ Map(αv)))
  def br0(test: Value, thn: => Ans, els: => Ans): Ans = ReaderTMonadPlus.mplus(thn, els)
\end{lstlisting}

>>>>

In this section, we instantiate @SPLattice@ with the flat binding-time
type @R[T] = T@ (i.e., @NoRep@); the result is the trait @Lattice[S]@.
An example of the lattices we use in the rest of the paper is the power set
abstract domain (shown in Figure \ref{fig:powerset}). Two power sets are
ordered by the subset relation. We use set union to compute their
join, and set intersection to compute their meet.  The bottom element of a
power set is the empty set, and we do not have a representation of the top element for
power sets.  Other lattices used in the paper, such as products and maps, can be
implemented similarly or by lifting the existing lattices element-wise or
point-wise.  Non-relational numerical abstract domains such intervals can also
be implemented in a stage-polymorphic way.

@@

In this section, we follow the abstracting definitional interpreters
(ADI) approach \cite{DBLP:journals/pacmpl/DaraisLNH17} to refactor our monadic
concrete interpreter into a monadic abstract interpreter.

@@

Similar to the concrete scenario, we sketch the @flatMap@ implementation of the
@SetT@ transformer and omit the rest. The field @run@ encapsulated by the monad
is of type @M[SetT[A]]@, where @M[_]@ is the inner monad, and @A@ is the element
type of the set. We first apply @flatMap@ on the inner monad to obtain the
set @s@. Then, we use the function @f@ to transform every element of type @A@
into a monadic value of type @Set[M, B]@. Finally, we fold all the transformed
values into a single monadic value of type @Set[M, B]@ by appending all of them.
The initial value of the fold is the empty value of @SetT@.
\begin{lstlisting}
  case class SetT[M[_]: Monad, A](run: M[Set[A]]) {
    def flatMap[B](f: A => SetT[M, B]): SetT[M, B] =
      SetT(Monad[M].flatMap(run) { (s: Set[A]) =>
        s.foldLeft(SetT.empty[M, B])((acc, a) => acc ++ f(a)).run
      }); ... }
\end{lstlisting}

@@

\paragraph{Primitive Operations} The primitive operations are changed according
to the new monad stack scheme and value domains. One of the notable changes due
to the nondeterminism is that the store-update operator merges the new values
with existing values for the same address in the store.  As we mentioned
before, sometimes we have to explore both of the branches for conditionals: in
method @br0@, we combine the results using the @mplus@ operation from
@MonadPlus@, which requires the value domains to be join-semilattices (in our
case, @Set@ and @Map@).
\begin{lstlisting}
  def set_store(αv: (Addr, Value)): AnsM[Unit] = liftM(StateTMonad.mod(σ => σ ⊔ Map(αv)))
  def br0(test: Value, thn: => Ans, els: => Ans): Ans = ReaderTMonadPlus.mplus(thn, els)
\end{lstlisting}
